---
title: "Measuring correlations"
subtitle: "Identifying empirical relationships"
author: "Austin Hart"
institute: "American University"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  message=FALSE, warning=FALSE, eval = TRUE, echo = FALSE, 
  fig.align = 'center', dev='svglite', fig.path = "images/"
)
```

```{r results='hide'}
library(tidyverse)
library(magrittr)
library(kableExtra)
library(janitor)

load('DCPS testing.rData')
load('bechdel.RData')
mov %<>%
  mutate(Era = if_else(year %in% 1970:1990, "1970-1990","1991-2013"))
```


# Identifying associations

### Three methods for today

- Cross-tabulation (categorical X, categorical Y)  

- Group comparison (categorical X, numeric Y)  

- Linear regression (numeric X, numeric Y)  

---
class: inverse, center, middle

# Recap
### Describing data on 1 variable


---
# Data vs dataset

- **DATA:** empirical information or observation

- **DATASET:** structured collection of data
  - Units `-` primary objects or entities under observation or study
  - Variables `-` measured characteristics of the units
  
- **VARIABLES:** What type of measure is that?
  - Categorical
  - Numeric
  
- Examples:
  1. [Data I use constantly](https://www.google.com/search?client=firefox-b-1-d&q=premier+league+table#sie=lg;/g/11pz7zbpnb;2;/m/02_tc;st;fp;1;;;)
  2. [Women in parliament](https://docs.google.com/spreadsheets/d/1sZkBcCiUvwX9JEmQhRiXg8SfEP--CxHlyJzoRNs-Hn4/edit#gid=993938300)



---
# Summarizing categorical data

### How many units fall into each group?

- "Bechdel test" data
- Evaluates 1,794 films 1970\-2013

```{r sumtab}
  tab = 
    mov %>%
    filter(Era == '1970-1990') %>%
    count(Bechdel) %>%
    mutate(
      Percent = 100 * n/sum(n)
    ) %>%
    adorn_totals()

  tab %>%
    kbl(digits = 1, format = "html", caption = "What women talk about, 1970-1990",
        col.names = c("Bechdel Score","N","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    column_spec(3, bold = T)

```

---
# Compare representation change over time

.pull-left[
```{r sumtabpre}
  tab = 
    mov %>%
    filter(Era == '1970-1990') %>%
    count(Bechdel) %>%
    mutate(
      Percent = 100 * n/sum(n)
    ) %>%
    adorn_totals()

  tab %>%
    kbl(digits = 1, format = "html", caption = "What women talk about, 1970-1990",
        col.names = c("Bechdel Score","N","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    column_spec(3, bold = T)

```
]

.pull-right[
```{r sumtabpost}
  tab = 
    mov %>%
    filter(Era != '1970-1990') %>%
    count(Bechdel) %>%
    mutate(
      Percent = 100 * n/sum(n)
    ) %>%
    adorn_totals()

  tab %>%
    kbl(digits = 1, format = "html", caption = "What women talk about, 1991-2013",
        col.names = c("Bechdel Score","N","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    column_spec(3, bold = T)

```
]

---
# Summarizing numeric data

### Basic statistics

- **Minimum**, $y_{min}$: lowest recorded score of $y$ in the data
- **Mean**, $\bar{y}$: simple average of recorded scores on $y$
- **Median**, $Q_2$: value of $y$ for the unit at the midpoint, or 50th percentile, of the ordered data.
- **Maximum**, $y_{max}$: highest recorded score of $y$


### Spread
- **Range**, $y_{max} - y_{min}$
- **Standard deviation**, $\sigma_y$: average distance of the cases from the mean
- **IQR**, $Q_3 - Q_1$: range capturing the middle 50% of observed data


---
# Summarize numeric data 

.pull-left[

&nbsp;

```{r d2desc}
mov %>%
  filter(Bechdel == 'ok') %>%
  summarise(
    Min = min(Budget,na.rm=T),
    Average = mean(Budget,na.rm=T),
    Median = median(Budget,na.rm=T),
    Max = max(Budget,na.rm=T),
    SD = sd(Budget,na.rm=T)
  ) %>%
  pivot_longer(1:5) %>%
  mutate(value = scales::dollar(value)) %>%
  kbl(digits = 1, format = "html", caption = "Bechdel Pass Films",
        col.names = c("","Budget, USD"), align = 'lr') %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center")
```
]

.pull-right[
```{r d2, fig.height=5.5, fig.width=4}

boxplot(mov$Budget[mov$Bechdel=='ok']/1000000,
     ylab = "USD, millions",
     ylim = c(0,500), col="#69b3a2",
     cex.lab=1.4, cex.axis=1.4, cex.sub=1.4,cex.main=1.8)

```
]

---
# Budgets in comparison

.pull-left[

&nbsp;

```{r d2desccomp}
mov %>%
  group_by(BechdelBinary) %>%
  summarise(
    Min = min(Budget,na.rm=T),
    Average = mean(Budget,na.rm=T),
    Median = median(Budget,na.rm=T),
    Max = max(Budget,na.rm=T),
    SD = sd(Budget,na.rm=T)
  ) %>%
  pivot_longer(2:6) %>%
  pivot_wider(names_from = BechdelBinary) %>%
  mutate_at(2:3, ~scales::dollar(.x)) %>%
  kbl(digits = 1, format = "html", caption = "Budget by Bechdel Test",
        align = 'lrr') %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center")
```
]

.pull-right[

```{r d2comp, fig.height=5.5, fig.width=4}

boxplot(mov$Budget/1000000 ~ mov$BechdelBinary,
     xlab = "Bechdel",
     ylab = NA,
     ylim = c(0,500), col="#69b3a2",
     cex.lab=1.4, cex.axis=1.4, cex.sub=1.4,cex.main=1.8)

```
]


---
class: inverse, center, middle

# MEASURING EMPIRICAL RELATIONSHIPS

---
# What is a correlation?

Given data on an exposure variable, $x$ and an outcome variable, $y$...

> *Statistical independence*: you find no pattern of association between $x$ and $y$. The distribution of $y$ is constant across different values or categories of $x$.

> *Correlation/dependence*: pattern of association between $x$ and $y$ in the data. The distribution of $y$ changes systematically "depending" on the value of $x$. 


---
# Correlation requires comparison

> Compare data on $y$ across different values or for different groups of $x$

.pull-left[
### Where is variation in $x$?
- Cross-sectional data
  - One observation/record per unit
  - $x$ differs across units
  
- Time-series data: 
  - Multiple records per unit
  - $x$ changes over time w/i units
]

.pull-right[
### Source of change in $x$?
- Experimental design
  - Researcher randomly assigns
  - Administers $x$  
  
- Observational design
  - Researcher observes $x$
  - "Nature" assigns/administers
]


---
class: center, middle

# Cross&ndash;tabulation

---
# Cross&ndash;tabulation

> Does the distribution on the outcome (i.e. relative frequency table) differ by groups?

- Assessing relationship between:  

  - Categorical exposure variable $(x)$  

  - Categorical outcome variable $(y)$

- Process: tabulate by group and compare!  


---
# You already did this!

.pull-left[
```{r sumtabpre2}
  tab = 
    mov %>%
    filter(Era == '1970-1990') %>%
    count(Bechdel) %>%
    mutate(
      Percent = 100 * n/sum(n)
    ) %>%
    adorn_totals()

  tab %>%
    kbl(digits = 1, format = "html", caption = "What women talk about, 1970-1990",
        col.names = c("Bechdel Score","N","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    column_spec(3, bold = T)

```
]

.pull-right[
```{r sumtabpost2}
  tab = 
    mov %>%
    filter(Era != '1970-1990') %>%
    count(Bechdel) %>%
    mutate(
      Percent = 100 * n/sum(n)
    ) %>%
    adorn_totals()

  tab %>%
    kbl(digits = 1, format = "html", caption = "What women talk about, 1991-2013",
        col.names = c("Bechdel Score","N","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    column_spec(3, bold = T)

```
]

---
# Putting it together

### A proper cross-tabulation

```{r xtabs}
  xtab =
    mov %>% 
    count(Bechdel,Era) %>%  
    na.omit() %>%
    pivot_wider(
      names_from = Era,
      values_from = n, 
      values_fill = 0
    ) %>%
    mutate_if(
      is.integer, list(Percent = ~./sum(.) * 100)
    ) %>% 
    select(Bechdel,sort(tidyselect::peek_vars())) %>%
    adorn_totals()

  xtab %>%
    kbl(digits = 1, format = "html", 
        caption = "Development of women characters over time", 
        col.names = c(" ","n","%","n","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    add_header_above(c(" " = 1,"1970-1990"=2,"1991-2013" = 2))
```

---
class: center, middle

# Group comparisons

---
# Comparing groups

Assessing the relationship between:  

- Categorical exposure variable $(x)$ 
- Continuous outcome variable $(y)$  

Process: Describe data in each group and compare!

---
# You've already done this too


.pull-left[

&nbsp;

```{r d2desccomp2}
mov %>%
  group_by(BechdelBinary) %>%
  summarise(
    Min = min(Budget,na.rm=T),
    Average = mean(Budget,na.rm=T),
    Median = median(Budget,na.rm=T),
    Max = max(Budget,na.rm=T),
    SD = sd(Budget,na.rm=T)
  ) %>%
  pivot_longer(2:6) %>%
  pivot_wider(names_from = BechdelBinary) %>%
  mutate_at(2:3, ~scales::dollar(.x)) %>%
  kbl(digits = 1, format = "html", caption = "Budget by Bechdel Test",
        align = 'lrr') %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center")
```
]

.pull-right[

```{r d2comp2, fig.height=5, fig.width=4}

boxplot(mov$Budget/1000000 ~ mov$BechdelBinary,
     xlab = "Bechdel",
     ylab = NA,
     ylim = c(0,500), col="#69b3a2",
     cex.lab=1.4, cex.axis=1.4, cex.sub=1.4,cex.main=1.8)

```
]

---
class: center, middle, inverse

# Linear regression


---
# The trends we see

### Looking for linear association, DCPS PARCC

```{r corr, fig.width=4, fig.height=4, dpi=250}
  # 2. plot  
    plot(ProfLang ~ ProfMath, data = dcps) # scatter

```

---
# Measuring linear correlation

### Simple linear dependence

- Correlation coefficient, $r_{xy}$: measures direction and degree of linear dependence $(-1 \leq r \leq 1)$.


### Models of the conditional mean

- Regression coefficient, $b_1$: estimated change in mean of $y$ for every one-unit increase in $x$


---
# Definining a conditional mean

Model $y$ as a linear function of $x$ and error: 
$$
y_i = b_0 + b_1(x_i) + e_i
$$

### Coefficients
- Intercept, $b_0 = E(Y|X=0)$
- Slope, $b_1 = E(\Delta Y | \Delta X)$

> The slope, $\beta$, captures the expected change in the mean of $y$ given a one-unit increase in $x$.


### The error term, $e_i$
- deviation from observed and predicted: $e_i = y_i - (b_0 + b_1(x_i))$
- reflects systematic bias and/or random error


---
# The data, the line, the error

From DCPS testing data: 
$$
ProfLang_i = 4.3 + 0.94(ProfMath_i) +e_i
$$

  A one-point increase in math proficiency is associated with a 0.94-point increase in mean language proficiency.
  

```{r fit1, fig.width=4, fig.height=4, dpi=250}
# Scatter w/OLS fit (numeric X, numeric Y)
  # 1. store OLS estimates
    est = lm(ProfLang ~ ProfMath, data = dcps)
  # 2. plot  
    plot(ProfLang ~ ProfMath, data = dcps) # scatter
    abline(est, col = 'cornflowerblue') # add linear fit
```
]


---
# What to do with your regression?

$$
ProfLang_i = 4.3 + 0.94(ProfMath_i) +e_i
$$

- Estimate average response in $y$ to unit-change in $x$

>   A one-point increase in math proficiency is associated with a 0.94-point increase in mean language proficiency. 

- Predict average $y$ for a given value of $x$ 
$$
\begin{aligned}
E(y_i|x_i=50) &= 4.3 + 0.94(50)\\
&= 4.3 + 47 \\
&= 51.3 \\
\end{aligned}
$$

---
# Precision of regression estimates

> Coefficient of Determination, $R^2$
> Captures percent of variation in y accounted for by x.

- Useful in comparing association across groups 
- Largely phasing out
  - Means very little in isolation
  - "prediction" is not typically the aim

---
class: inverse

# Mandatory warning labels

### Is this causal?
- Correlation is a necessary but insufficient condition for causal inference.
- All/some/none of the relationship you identify may be evidence of the causal effect... 
  - of $x$ on $y$
  - of $z$, an omitted cause of $x$ and $y$

### Is this systematic?
- Could this be the result of chance alone?
- What does it say about a broader population of cases?

